---
title: "IDA_adnimerge"
author: "Yibin Feng"
date: "1/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install ADNIMERGE

```{r}
#install.packages("Hmisc")
#install.packages("restricted/ADNIMERGE_0.0.1.tar.gz", repos=NULL, type="source")
```

# Load package

```{r}
rm(list = ls())

library(tidyverse)
library(ggpubr)
library(ggfortify)


library(Hmisc)
library(ADNIMERGE)
```


# Instruction

- ADNIMERGE package contains all ADNI data files
- One of the dataframes is adnimerge which contains commonly used variables.


```{r, eval=FALSE}
help(package = "ADNIMERGE")
```

# Utility function
Define some utility functions for future use here.

```{r}
get_datadic <- function(variables, table = 'ADNIMERGE', all = F){
  # Retrieve all info from datadict, or most frequently used info
  if (all) {
    subset(datadic, TBLNAME == table & FLDNAME %in% variables)
  } else {
    subset(datadic, TBLNAME == table & FLDNAME %in% variables,
           c('FLDNAME', 'TBLNAME', 'TEXT', 'NOTES'))
  }
}
```

# IDA of dataframe `adnimerge`

This section does not involve any data preparation.

## Variable names

List all
```{r, eval=FALSE}
names(adnimerge)
```

variable names containing ".bl", note `Years.bl` and `Month.bl` are not baseline values but time-varying
```{r}
names(adnimerge)[grepl(".bl", names(adnimerge))]
```
variable names not containing ".bl"
```{r}
names(adnimerge)[!grepl(".bl", names(adnimerge))]
```

Baseline covariates
```{r}
vars_baseline <- c("AGE", "PTGENDER", "PTEDUCAT", "PTETHCAT", "PTRACCAT", "PTMARRY", "APOE4")

get_datadic(vars_baseline)
```


## Identifier for subject

Look up the datadic.
```{r}
get_datadic(variables = "RID")
```

Count unique RIDs.
```{r}
print(paste("Number of subjects in adnimerge =", length(unique(adnimerge$RID))))
```

## Multiple identifier for time
There are multiple variables related to time i.e. "VISCODE", "EXAMDATE", "Years.bl", "Month.bl", "Month", "M".
What are the differences?
Look up datadic
```{r}
get_datadic(variables = c("VISCODE", "EXAMDATE", "Years.bl", "Month.bl", "Month", "M"))
```

We summarize the findings here and elaborate them further below:
- `Month`, `M` are not based on exact date.
- `Month` is factor, but `M` is numeric (integer).
- `Month` may differs from `M`, despite both are said to be months since baseline.
- `M` is mostly consistent with `VISCODE` e.g. M==6 and VISCODE==m06.
- `Years.bl`, `Month.bl` description are missing in datadic.
- `Years.bl`, `Month.bl` do not represent a baseline value and are time-varying, contrary to other variables with `.bl` suffix e.g. `EXAMDATE.bl`.
- `Years.bl`, `Month.bl` measures the actual follow-up time (in years or months) after baseline `EXAMDATE`.


```{r}
adnimerge %>%
  select("VISCODE", "Month", "M", "EXAMDATE", "EXAMDATE.bl", "Years.bl", "Month.bl") %>%
  str()
```

"M" is consistent with the "VISCODE"
```{r}
tmp_viscode <- adnimerge$VISCODE
tmp_viscode[tmp_viscode == "bl"] <- "m00" # Set bl to m00, so format is consistent with m06, m12 etc.
tmp_viscode <- sapply(tmp_viscode, function(x) {
  as.numeric(strsplit(x, split = "m")[[1]][2])
})

print(paste("[Check] all VISCODE identical to M:", all(tmp_viscode == adnimerge$M)))
```

Construct the Years.bl and Month.bl using EXAMDATE:
```{r}
adnimerge %>%
  filter(RID == 2) %>%
  select("VISCODE", "Month", "M", "EXAMDATE", "EXAMDATE.bl", "Years.bl", "Month.bl") %>%
  mutate(verify_Days.bl = as.numeric(EXAMDATE - EXAMDATE.bl),
         verify_Years.bl = verify_Days.bl / 365.25,
         verify_Months.bl = verify_Years.bl * 12)
```

Now we compute the differences using our deduced conversion formula:

Years.bl = (EXAMDATE - EXAMDATE.bl) / 365.25?
```{r}
# Deduced that two should be convertible from Month.bl = (EXAMDATE - EXAMDATE.bl) / 366 * 12
adnimerge %>% 
  select(RID, EXAMDATE.bl, EXAMDATE, Years.bl) %>%
  mutate(time_followup = as.numeric((EXAMDATE - EXAMDATE.bl) / 365.25)) %>%
  mutate(abs_diff = abs(time_followup - Years.bl)) %>%
  select(abs_diff) %>% colSums()
```

Month.bl = (EXAMDATE - EXAMDATE.bl) / 365.25 * 12?
```{r}
# Deduced that two should be convertible from Month.bl = (EXAMDATE - EXAMDATE.bl) / 366 * 12
adnimerge %>% 
  select(RID, EXAMDATE.bl, EXAMDATE, Month.bl) %>%
  mutate(time_followup = as.numeric((EXAMDATE - EXAMDATE.bl) / 365.25 * 12)) %>%
  mutate(abs_diff = abs(time_followup - Month.bl)) %>%
  select(abs_diff) %>% colSums()
```

One more thing, are `M` and `Month` identical?
Not always!
```{r}
(adnimerge$Month == adnimerge$M) %>% 
  table()
```

Contrast the differences between `M` and `Month`:

levels in `Month`:
```{r}
as.numeric(levels(adnimerge$Month))
```

unique values of `M`:
```{r}
sort(unique(adnimerge$M))
```

When does these two disagree?
For example, we observe that for subject RID 31, `Month` == 60 and 72 appear twice.
We consider the `Month.bl` equivalent to `EXAMDATE` as the ground truth.
The first `Month` == 60 matches with the `Month.bl` exactly, but there is mismatch with `VISCODE` == 54.
The second `Month` == 60 differs from `Month.bl` by 1 month, and matches the `VISCODE`.

A guess is that `Month` is factorize on a predefined, regular grid, and the actual visit time is approximated to the nearest grid.
There is a risk double entry at one time may cause data error in some modeling using regular time grid, e.g. MFPCCox

```{r}
adnimerge %>%
  select(RID, VISCODE, EXAMDATE, Month.bl, Month, M) %>%
  mutate(is_difference = !(Month == M)) %>%
  filter(RID == 31, M > 42, M < 84) # Pick subject RID 31 as example
```

Look at data gathered at two `Month` == 60 took place on 2010-10-28 and 2010-12-01:
- From two different COLPROT (collection protocol) ADNI1 and ADNIGO
- No measurement values for first visit.

```{r}
adnimerge %>%
  filter(RID == 31, Month == 60) %>%
  select(names(adnimerge)[!grepl(".bl", names(adnimerge))]) # Exclude .bl
```

Similar at two `Month` == 72:
- From two different COLPROT (collection protocol) ADNIGO and ADNI2
- No measurement values for first visit.

```{r}
adnimerge %>%
  filter(RID == 31, Month == 72) %>%
  select(names(adnimerge)[!grepl(".bl", names(adnimerge))]) # Exclude .bl
```

List all subjects with this `Month` != `M` issue to have a broader look:
```{r}
adnimerge %>%
  select(RID, VISCODE, EXAMDATE, Month.bl, Month, M) %>%
  filter(!(Month == M))
```

Subset potentially problematic observations with this problem to probe into observations:
- Not all cases related to changing COLPROT, see RID 2
- Not all case involve double entry, see RID 29

### Considerations for data cleaning

Two problems we found when selecting the time label, first is double entry, second is mismatch between M, Month, Month.bl

1) whether double entry will happen on `Month` or `M`?'
Both.
In all cases except one, the `Month` count is less than number of visits, implying double entry.
In one case, RID 6014, the `M` count is less than number of visits, n_visit = 4, n_M = 3

2) which one is closer to the ground truth `EXAMDATE` when used as time label?
`Month` by eye test

3) at times of double entry, whether these entries contain different values or single value?
RID 98, `Month` == 12 has two entries, contain different values.
RID 6014, `M` == 0 has two entries, only one entry contains values, other one is NAs.

3) if we have to use a regular time grid for some methods, whether to use `Month` or `M`?
`M` (numeric, stick to viscode, more different from EXAMDATE) or `Month` (closer to truth, need to handle different values at same time label, factor)

```{r}
tmp <- adnimerge %>%
  select(RID, VISCODE, EXAMDATE, Month.bl, Month, M) %>%
  filter(!(Month == M)) %>%
  select(RID, Month)

tmp2 <- lapply(1:nrow(tmp), function(i) {
  adnimerge %>%
    filter(RID == tmp[i, "RID"], Month == tmp[i, "Month"]) %>%
    select(Month.bl, names(adnimerge)[!grepl(".bl", names(adnimerge))]) # Exclude .bl
})

tmp3 <- do.call(rbind, tmp2) %>%
  select(RID, COLPROT, ORIGPROT, VISCODE, M, Month, Month.bl, EXAMDATE, everything())

tmp3
```

descriptive statistics of this issue:
```{r}
adnimerge %>%
  #filter(RID == 31) %>%
  group_by(RID) %>%
  dplyr::summarize(
    n_visit = n(),
    n_Month = length(unique(Month)),
    n_M = length(unique(M)),
    visit_Month = n_visit == n_Month,
    visit_M = n_visit == n_M, 
    all_n_equal = visit_Month & visit_M
    ) %>%
  filter(!all_n_equal)
```

Probe RID 6014, we find that:
- `M` == 0 appears twice and `EXAMDATE` are close
- `Month` == 3 is way off and incorrect to describe EXAMDATE
- good news is that we can drop the one filled with NAs `Month` == 3
- but no idea why it exists at first place?
```{r}
adnimerge %>%
  filter(RID == 6014) %>%
  select(names(adnimerge)[!grepl(".bl", names(adnimerge))]) %>% # Exclude .bl
  select(Month, M, VISCODE, EXAMDATE, everything())
```
## Ethnicity and race

Underrepresented. Make sure to mention this in thesis.
Why white is dominant in this population?
What is the difference between ethnicity Hisp/Latino and other?

```{r}
adnimerge %>% 
  select(PTETHCAT, PTRACCAT) %>% 
  summary()
```



## Age

- There are NAs in baseline age, need to be cleaned
```{r}
adnimerge %>%
  select(AGE) %>%
  summary()
```

## Follow up period

```{r}
adnimerge %>%
  group_by(RID) %>%
  dplyr::summarize(last.visit = max(Years.bl),
            ORIGPROT = unique(ORIGPROT)) %>%
  ggplot() + 
    geom_histogram(aes(last.visit), 
                   binwidth = 1) +
    facet_wrap(vars(ORIGPROT)) +
    labs(title = "Observation period in each ADNI phase",
         x = "Years since baseline")
```

```{r}
adnimerge %>%
  group_by(RID) %>%
  dplyr::summarize(last.visit = max(Years.bl),
            ORIGPROT = unique(ORIGPROT)) %>%
  ggplot() + 
    geom_histogram(aes(last.visit), 
                   binwidth = 1) +
    facet_wrap(vars(ORIGPROT)) +
    labs(title = "Observation period by each ADNI phase",
         x = "Years since baseline")
```

```{r}
adnimerge %>%
  group_by(RID) %>%
  dplyr::summarize(last.visit = max(Years.bl),
            ORIGPROT = unique(ORIGPROT),
            DX.bl = unique(DX.bl)) %>%
  ggplot() + 
    geom_jitter(aes(last.visit, DX.bl,
                    color = DX.bl)) +
    facet_wrap(vars(ORIGPROT)) +
    labs(title = "Observation period by baseline diagnosis DX.bl in different ADNI phase",
         x = "Year since baseline")
```

### Consideration for data cleaning
Exclude subjects DX == NA or Dementia at baseline

```{r}
adnimerge %>%
  group_by(RID) %>%
  dplyr::summarize(last.visit = max(Years.bl),
            ORIGPROT = unique(ORIGPROT)) %>%
  left_join(adnimerge %>% filter(VISCODE == "bl") %>% select(RID, DX), by = "RID") %>%
  ggplot() + 
    geom_jitter(aes(last.visit, DX,
                    color = DX),
                alpha = 0.5) +
    facet_wrap(vars(ORIGPROT)) +
    labs(title = "Observation period by diagnosis DX at baseline in different ADNI phase",
         subtitle = "baseline filter by VISCODE = bl",
         x = "Years since baseline")
```

```{r}
# distribution of follow-up period
adnimerge %>% 
  group_by(RID) %>%
  dplyr::summarize(fup.period = max(Years.bl)) %>%
  select(fup.period) %>%
  summary()

```

## Average distinct values per variables

```{r, eval=FALSE}
# Compute average distinct values per variables
adnimerge %>%
  group_by(RID) %>%
  select(names(adnimerge)[!grepl(".bl", names(adnimerge))]) %>% # Exclude .bl
  dplyr::summarise(across(everything(), n_distinct)) %>%
  dplyr::select(-RID) %>%
  dplyr::summarise(across(everything(), mean)) %>%
  pivot_longer(cols = everything(), names_to = "var_name", values_to = "avg_distinct_values") %>%
  arrange(desc(avg_distinct_values)) %>% 
  View()
```

```{r, eval=FALSE}
# Compute max distinct values per variables
adnimerge %>%
  group_by(RID) %>%
  select(names(adnimerge)[!grepl(".bl", names(adnimerge))]) %>% # Exclude .bl
  dplyr::summarise(across(everything(), n_distinct)) %>%
  dplyr::select(-RID) %>%
  dplyr::summarise(across(everything(), max)) %>%
  pivot_longer(cols = everything(), names_to = "var_name", values_to = "max_distinct_values") %>%
  arrange(desc(max_distinct_values)) %>% 
  View()
```

## Number of visits

### Consideration for data cleaning
Exclude subjects with only single visit i.e. at baseline only, no repeated measurement

```{r}
ids_one_visit <- adnimerge %>%
  group_by(RID) %>%
  dplyr::summarize(last.visit = max(M)) %>%
  filter(last.visit == 0) %>%
  select(RID) %>%
  unlist()

adnimerge %>%
  filter(RID %in% ids_one_visit) %>%
  select(RID, DX.bl, DX)
```


```{r}
adnimerge %>% 
  group_by(RID) %>%
  dplyr::summarize(n.visits = n()) %>%
  ggplot() +
    geom_bar(aes(n.visits)) +
    labs(
      title = "Number of visits",
      subtitle = "But not every visit has measurements, observation is irregular!"
    )
```

## NAs in baseline covariates
```{r}
vars_baseline <- c("AGE", "PTGENDER", "PTEDUCAT", "PTETHCAT", "PTRACCAT", "PTMARRY", "APOE4")
adnimerge %>% 
  select(vars_baseline) %>% 
  summary()
```


## Event of interest i.e. diagnosis
```{r, eval=FALSE}
# Check DX at baseline
adnimerge %>% filter(VISCODE == "bl") %>% select(DX) %>% summarytools::freq()
```

### Who are the DX.bl == "AD"?

```{r}
adnimerge %>% 
  filter(DX.bl == "AD") %>%
  select(RID, DX.bl, DX, EXAMDATE, M) %>%
  tail()
```

```{r}
# At baseline row
with(adnimerge %>% 
       filter(VISCODE=="bl"), summarytools::ctable(DX.bl, DX))
```


```{r}
# DX.bl == AD will sometimes lead to DX == MCI rather then Dementia
with(adnimerge, summarytools::ctable(DX.bl, DX))
```


```{r}
# Further investigate
# Participant 162, 167, 739, 4641, 4845
adnimerge %>%
  filter(DX.bl == "AD", DX == "MCI") %>%
  select(RID, EXAMDATE, M, DX.bl, DX)
```


```{r}
# Reversion seems to be rare but possible according to ADNI slide?
adnimerge %>%
  filter(RID %in% c(162, 167, 739, 4641, 4845)) %>%
  select(RID, EXAMDATE, M, DX.bl, DX) %>%
  ggplot() + 
    geom_point(aes(M, DX, color = DX)) + 
    labs(title = "PTs with AD at baseline but with MCI instead of Dementia at follow up",
         caption = "Baseline refers to DX.bl, follow up refers to DX",
         x = "Month since baseline") + 
    facet_wrap(vars(RID)) + 
    theme(legend.position = "top")
```

# Check dropout 

```{r}
# Early Discontinuation and Withdrawal
#ADNIMERGE::treatdis %>% head() %>% View()

ADNIMERGE::treatdis %>% 
  filter(RID %in% df.surv_preds$RID) %>% 
  select(WDREASON) %>% 
  table()
```


```{r}
(df.surv_preds$RID %in% ADNIMERGE::treatdis$RID) %>% 
  mean() # Not all subjects in cleaned data found in Early Discontinuation and Withdrawal
# Likely reason: not drop out yet? new RIDs not included => up to ADNI2

ADNIMERGE::treatdis %>% 
  filter(RID %in% df.surv_preds$RID) %>% 
  nrow() # Check n size

# Count death frequency
sapply(
  ADNIMERGE::treatdis %>% 
    filter(RID %in% df.surv_preds$RID) %>% 
    select(WDREASON), 
  function(reason){
   grepl("Death", reason)
  }) %>% sum()
```


