---
title: "Untitled"
author: "Yibin Feng"
date: '2022-07-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Purpose
This markdown is for extracting folds.eval and visualize the results in tables and figures.

# Dependencies
```{r}
# Dependencies
library(tidyverse)
library(ggrepel)

source("function_utility.R")
source("function_evaluation.R")
```

# Load results (folds.eval) to vector of names and paths from directory
```{r}
# Where the folds.eval are placed
#folderpath <- "./output/comparison/" # copy folds.eval to here for selected comparison
folderpath <- "./output/" # Immediate output location of folds.eval
n_fold <- 10
deltaT <- 1:15

models.path <- list.files(path = folderpath) # Using repeated measurements up to landmark time

# Filter folds.eval files only
models.path <- models.path[grepl(pattern = "eval_", x = models.path)]

# Filter by landmark
#models.path <- models.path[grepl(pattern = "lm2", x = models.path) | grepl(pattern = "lm0", x = models.path)] # Chose landmark

# Extract model names
models.name <- stringr::str_remove_all(stringr::str_remove_all(models.path, ".RData"), "eval_")

# Check number of models loaded
print(paste("Number of folds.eval available:", length(models.name)))
```

# tdAUC
## Consolidate
```{r}
# --------------------------------------------------
# Summarize tdAUC results
# --------------------------------------------------
Combine_res_tdauc <- function(
    models.name, models.path, T.start, deltaT
) {
  # Initialize
  df.tdauc.compare <- data.frame()

  deltaT <- deltaT[deltaT > T.start]
  
  is_lm <- grepl(models.name, pattern = paste0("lm", T.start))
  models.name.valid <- models.name[is_lm]
  models.path.valid <- models.path[is_lm]
# --------------------------------------------------
  
  # Repeat the following steps for different models
  
  for (i in 1:length(models.name.valid)) {
    
    name <- models.name.valid[i]
    path <- models.path.valid[i]
      
    if (i == 1) {
      # For first model to initialize dataframe
      df.tdauc <- Summarize.tdauc(
        name = name,
        path = paste0(folderpath, path),
        deltaT = deltaT
        )
      
      df.tdauc.compare <- df.tdauc
    } else {  #  i > 1
      df.tdauc <- Summarize.tdauc(
        name = name,
        path = paste0(folderpath, path),
        deltaT = deltaT
      )
    
      # Combine results from other models to compare dataframe
      df.tdauc.compare <- df.tdauc.compare %>%
        add_row(df.tdauc)
    }
  }
  return(df.tdauc.compare)
}
# --------------------------------------------------

# Example
# T.start <- 2
# deltaT <- 1:15
# df.tdauc.compare <- Combine_res_tdauc(models.name, models.path, T.start, deltaT)
# dim(df.tdauc.compare)
```

```{r}
# --------------------------------------------------
# Summarize over all landmark times

list.tdauc.lms <- lapply(2:6, function(T.start) {
  df.tdauc.compare <- Combine_res_tdauc(models.name, models.path, T.start, deltaT)
  
  return(list(
    landmark = T.start,
    df.tdauc.compare = df.tdauc.compare
  ))
})
# --------------------------------------------------
# Check
for (x in list.tdauc.lms) {
  x$df.tdauc.compare %>%
    group_by(method, model.name) %>%
    summarize(n(), .groups = "drop") %>%
    print()
}
```

## Fig: tdAUC | LM (WARNING: group by method, not method.full)

```{r}
g_title <- "Comparison of estimated Brier score under repeated 10-fold CV"
g_caption <- paste0(
  "CV repeated with 10 different seeds."
  )
labels.method <- c("MFPCCox", "pCox-baseline", "pCox-landmarking", "PRC-LMM")
# --------------------------------------------------
# Create plots
list.g.tdauc <- lapply(list.tdauc.lms, function(x) {
  df <- x$df.tdauc.compare
  t_landmark <- x$landmark
  
  g <- df %>% 
  group_by(prediction_time, model.name) %>%
  summarize(repeated.CV.mean = mean(mean),
            repeated.CV.max = max(mean),
            repeated.CV.min = min(mean),
            method = unique(method),
            method.full = unique(method.full),
            n_bl.covariate = unique(n_bl.covariate),
            .groups = "drop") %>%
#  mutate(method.and.n_bl.cov = paste(method, n_bl.covariate, sep = "_")) %>%
  ggplot(aes(
    x = prediction_time, 
    y = repeated.CV.mean, 
    group = method,
    color = method
)
  ) +
 # geom_errorbar(aes(
 #   ymax = repeated.CV.max,
 #   ymin = repeated.CV.min),
 #   alpha = 0.3,
 #   width = 0.3, size = 0.8) +
  # landmark
  geom_vline(xintercept = t_landmark, 
             color = "grey80", linetype = "dashed", size = 1, alpha = 1) + 
  geom_line(
    aes(linetype = method), 
    alpha = 0.8, size = 0.8
    ) +
  geom_point(
    aes(shape = method), 
    alpha = 0.8, size = 2.5
    ) +
  scale_x_continuous(breaks = 1:15) + 
  scale_y_continuous(breaks = 0:10 / 10) +
  coord_cartesian(
    xlim = c(0, 15),
    ylim = c(0.7, 1)
    ) +
  ggsci::scale_color_lancet(labels = labels.method) +
  scale_linetype_discrete(labels = labels.method) +
  scale_shape_discrete(labels = labels.method) +
  xlab("Year from baseline") + ylab("tdAUC") +
  labs(
#    title = g_title,
    title = paste("Landmark time =", t_landmark),
#    caption = g_caption,
#    caption = "Error bar represents the max and min",
#    linetype = "number of (candidate) baseline covariates",
    color = "Method",
    linetype = "Method",
    shape = "Method"
  )  +
  theme_pubr() +
  theme(
    legend.position = "right", 
    legend.box = "vertical",
    panel.grid.major.y = element_line(color = "grey60"),
    panel.grid.minor.y = element_line(color = "grey80")
    ) 
  return(list(
    landmark = t_landmark,
    g = g
  ))
})
```

print all tdAUC
```{r, fig.width=12, fig.height=6}
# Print
for (x in list.g.tdauc) {
  print(x$g)
}
```

save all tdAUC
```{r}
for (x in list.g.tdauc) {
  ggsave(
    filename = paste0("compare_tdauc_lm", x$landmark, ".png"),
    plot = x$g,
    device = "png",
    path = "./figure/compare_tdauc/",
    width = 12, height = 6
  )
}

```

combined plot
```{r, fig.width=8, fig.height=12}
g.tdauc.combined <- ggpubr::ggarrange(
  plotlist = lapply(list.g.tdauc, function(x) x$g), 
  ncol = 2, nrow = 3, common.legend = TRUE
)

ggsave(
  filename = "compare_tdauc_combined.png",
  plot = g.tdauc.combined,
  device = "png", 
  path = "./figure/compare_tdauc/",
  width = 10, height = 12,
  bg = "white"
)
```


## Draft: Compare all tdAUC in all LMs | MFPCCox only

```{r}
# --------------------------------------------------
# Create plots
list.g.tdauc <- lapply(list.tdauc.lms, function(x) {
  df <- x$df.tdauc.compare
  t_landmark <- x$landmark
  
  g <- df %>%
  filter(method == "MFPCCox") %>%
  group_by(prediction_time, model.name) %>%
  summarize(repeated.CV.mean = mean(mean),
            repeated.CV.max = max(mean),
            repeated.CV.min = min(mean),
            method = unique(method),
            method.full = unique(method.full),
            n_bl.covariate = unique(n_bl.covariate),
            .groups = "drop") %>%
#  mutate(method.and.n_bl.cov = paste(method, n_bl.covariate, sep = "_")) %>%
  ggplot(aes(
    x = prediction_time, 
    y = repeated.CV.mean, 
    group = method.full,
    color = method.full
)
  ) +
 # geom_errorbar(aes(
 #   ymax = repeated.CV.max,
 #   ymin = repeated.CV.min),
 #   alpha = 0.3,
 #   width = 0.3, size = 0.8) +
  geom_vline(xintercept = t_landmark, 
             color = "grey80", linetype = "dashed", size = 1, alpha = 1) + # landmark
  geom_line(
    aes(linetype = method.full), 
    alpha = 0.8, size = 0.8
    ) +
  geom_point(
    aes(shape = method.full), 
    alpha = 0.8, size = 2.5
    ) +
  scale_x_continuous(breaks = 1:15) + 
  scale_y_continuous(breaks = 0:10 / 10) +
  coord_cartesian(
    xlim = c(0, 15),
    ylim = c(0.7, 1)
    ) +
  ggsci::scale_color_lancet() +
  xlab("Year from baseline") + ylab("tdAUC") +
  labs(
    title = "Comparison of estimated tdAUC under repeated 10-fold CV",
    subtitle = paste("at landmark time", t_landmark),
    caption = "CV repeated in 10 different seeds",
#    caption = "Error bar represents the max and min",
#    linetype = "number of (candidate) baseline covariates",
    color = "Method",
    shape = "Method",
    linetype = "Method"
  )  +
  theme_pubr() +
  theme(
    legend.position = "right", 
    legend.box = "vertical",
    panel.grid.major.y = element_line(color = "grey60")
    ) 
  return(list(
    g = g
  ))
})
```

```{r, fig.width=12, fig.height=6}
# Print
for (x in list.g.tdauc) {
  print(x$g)
}
```


# C index 

## Consolidate
```{r}
# --------------------------------------------------
# Summarize C-index results
# --------------------------------------------------

# Initialize
df.c.index.compare <- data.frame()
# ------------------------------------------------------------------------------------------------

for (i in 1:length(models.name)) {
  
  if (i == 1) {
    # Result 1
    df.c.index <- Summarize.c.index(
      model = models.name[i], 
      path = paste0(folderpath, models.path[i])
      )
    
    df.c.index.compare <- df.c.index
  } else {
    # Result i, i > 1
    df.c.index <- Summarize.c.index(
      model = models.name[i], 
      path = paste0(folderpath, models.path[i])
      )
    
    # Combine results from other models to compare dataframe
    df.c.index.compare <- df.c.index.compare %>%
      add_row(df.c.index)  
  }
}
# --------------------------------------------------
# Check
df.c.index.compare %>%
    group_by(method, method.full) %>%
    summarize(n(), .groups = "drop") %>%
    print()
```

## Figure: C index | LM (group by method)
group by method is not method.full
```{r}
# --------------------------------------------------
# Plot df.c.index.compare
# --------------------------------------------------
g_caption <- paste0(
  "CV repeated with 10 different seeds.\n",
  "Box plot refer to the IQR (from first to third percentile).\n",
  "Whiskers extend from 1.5*IQR for either end of the IQR box."
  )

g_title <- "Comparison of estimated C-index under repeated 10-fold CV"

labels.method <- c("MFPCCox", "pCox-baseline", "pCox-landmarking", "PRC-LMM")
# --------------------------------------------------
# https://ggplot2.tidyverse.org/reference/geom_boxplot.html
list.g.c <- lapply(2:6, function(t_landmark) {
    g <- df.c.index.compare %>%
    filter(landmark == paste0("lm", t_landmark)) %>%
    # filter(grepl(model, pattern = "scaled")) %>% # dumb method to filter based on method.full
    # filter(grepl(model, pattern = "scenario2")) %>%
  #  filter(scenario == "scenario2" | scenario == "scenario0") %>%
  #  filter(method == "pCox-lm" | method == "pencal" | method == "pCox-bl" | method == "pencal-min" | method == "pencal-1se") %>%
    ggplot(aes(
      x = method, 
      y = c.index, 
      fill = method,
      color = method)) + 
      geom_point(
        aes(shape = method),
        position = position_jitter(seed = 1, width = 0.2),
        alpha = 0.2, 
        size = 1
        ) +
      geom_violin(
        color = NA,
        width = 0.5, 
        alpha = 0.2
        ) + 
      geom_boxplot(
        fill = "white",
        lwd = 0.8,
        width = 0.1,
        alpha = 1,
        outlier.shape = NA
        ) + 
      coord_cartesian(ylim = c(0.5, 1)) +
      ggsci::scale_color_lancet() +
      ggsci::scale_fill_lancet() +
      xlab("") + ylab("C-index") +
      labs(
        title = paste("Landmark time =", t_landmark),
#        caption = g_caption,
        color = "Method",
        fill = "Method",
        shape = "Method"
        ) + 
      theme_pubr() +
      theme(
        legend.position = "right",
        legend.box = "vertical",
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey60"),
        panel.grid.minor.y = element_line(color = "grey80")
      )
    
    return(g)
})
```

Inspect per landmark time
```{r, fig.width=12, fig.height=6}
list.g.c[[1]]
```


```{r, fig.width=16, fig.height=8}
g.c.wrapped <- ggpubr::ggarrange(
  plotlist = list.g.c, 
  ncol = 3, nrow = 2, 
  common.legend = TRUE
)

g.c.wrapped

ggsave(
    filename = paste0("compare_c_wrapped.png"),
    plot = g.c.wrapped,
    device = "png",
    path = "./figure/compare_c/",
    width = 16, height = 8,
    bg = "white"
)
```

## Figure: combined C index | LM (group by method)
```{r, fig.width=16, fig.height=8}
g_caption <- paste0(
  "CV repeated with 10 different seeds.\n",
  "Box plot refer to the IQR (from first to third percentile).\n",
  "Whiskers extend from 1.5*IQR for either end of the IQR box."
  )

g_title <- "Comparison of estimated C-index under repeated 10-fold CV"
labels.method <- c("MFPCCox", "pCox-baseline", "pCox-landmarking", "PRC-LMM")
# --------------------------------------------------
g.c.combined <- df.c.index.compare %>%
  mutate(landmark = as.numeric(
    stringr::str_split(landmark, pattern = "lm", simplify =  TRUE)[, 2])
    ) %>%
  ggplot(aes(
      x = method, 
      y = c.index, 
      group = interaction(landmark, method),
      fill = method,
      color = method)) + 
      geom_violin(
        color = NA,
        width = 0.8, 
        alpha = 0.3
        ) + 
      geom_boxplot(
        fill = NA, color = "black",
        lwd = 0.5,
        width = 0.2, 
        position = position_dodge(width = 0.8),
        alpha = 1,
        outlier.shape = NA
        ) + 
      coord_cartesian(ylim = c(0.5, 1)) +
      ggsci::scale_color_lancet(labels = labels.method) +
      ggsci::scale_fill_lancet(labels = labels.method) +
      scale_x_discrete(labels = labels.method) +
      xlab("") + ylab("C-index") +
      labs(
#        title = "Comparison of estimated C-index under repeated 10-fold CV",
#        subtitle = "at landmark times 2 (left) to 6 (right)",
#        caption = g_caption,
        color = "Method",
        fill = "Method"
        ) + 
      theme_pubr() +
      theme(
        legend.position = "top",
        legend.box = "vertical",
        axis.ticks.x = element_blank(),
#        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey60"),
        panel.grid.minor.y = element_line(color = "grey80")
      )

g.c.combined

ggsave(
    filename = paste0("compare_c_combined.png"),
    plot = g.c.combined,
    device = "png",
    path = "./figure/compare_c/",
    width = 12, height = 4,
    bg = "white"
)
```

## Table: C index
```{r}
table.c.index <- df.c.index.compare %>%
  group_by(landmark, method) %>%
  summarise(`C index` = round(mean(c.index, na.rm = TRUE), 3), .groups = "drop") %>%
  pivot_wider(names_from = landmark, values_from = `C index`)

table.c.index

Hmisc::latex(
  object = table.c.index,
  title = "C index",
  label = "tab:c.index",
  rowlabel = NULL,
  booktabs = TRUE
)


```




# Brier score

## Consolidate
```{r}
# --------------------------------------------------
# Summarize Brier score results
# --------------------------------------------------
Combine_res_brier <- function(
    models.name, models.path, T.start, deltaT
) {
  # Initialize
  df.brier.compare <- data.frame()
# --------------------------------------------------
  deltaT <- deltaT[deltaT > T.start]
  
  is_lm <- grepl(models.name, pattern = paste0("lm", T.start))
  models.name.valid <- models.name[is_lm]
  models.path.valid <- models.path[is_lm]
# --------------------------------------------------
  
  # Repeat the following steps for different models
  
  for (i in 1:length(models.name.valid)) {
    
    name <- models.name.valid[i]
    path <- models.path.valid[i]
      
    if (i == 1) {
      # For first model to initialize dataframe
      df.brier <- Summarize.brier(
        name = name,
        path = paste0(folderpath, path),
        deltaT = deltaT
        )
      
      df.brier.compare <- df.brier
    } else {
      # Result i, i > 1
      df.brier <- Summarize.brier(
        name = name,
        path = paste0(folderpath, path),
        deltaT = deltaT
        )
      
      # Combine results from other models to compare dataframe
      df.brier.compare <- df.brier.compare %>%
        add_row(df.brier)
    }
  }
  return(df.brier.compare)
}
# --------------------------------------------------
# Example
# T.start <- 2
# deltaT <- 1:15
# 
# df.brier.compare <- Combine_res_brier(models.name, models.path, T.start, deltaT)
# 
# dim(df.brier.compare)
```

```{r}
# --------------------------------------------------
# Summarize over all landmark times
list.brier.lms <- lapply(2:6, function(i) {
  T.start <- i
  df.brier.compare <- Combine_res_brier(models.name, models.path, T.start, deltaT)
  
  return(list(
    landmark = T.start,
    df.brier.compare = df.brier.compare
  ))
})
# --------------------------------------------------
# Check
# Check
for (x in list.brier.lms) {
  x$df.brier.compare %>%
    group_by(method, method.full) %>%
    summarize(n(), .groups = "drop") %>%
    print()
}
```

## Fig: BS | LM (group by method)
```{r}
g_title <- "Comparison of estimated Brier score under repeated 10-fold CV"
g_caption <- paste0(
  "CV repeated with 10 different seeds."
  )
labels.method <- c("MFPCCox", "pCox-baseline", "pCox-landmarking", "PRC-LMM")
# --------------------------------------------------
# Create plots
list.g.brier <- lapply(list.brier.lms, function(x) {
# --------------------------------------------------
  # Plot df.tdauc.compare
# --------------------------------------------------
  df <- x$df.brier.compare
  t_landmark <- x$landmark
  
  g <- df %>% 
  #  filter(scenario == "scenario2" | scenario == "scenario0") %>%
    group_by(prediction_time, model.name) %>%
    summarize(repeated.CV.mean = mean(mean),
              repeated.CV.max = max(mean),
              repeated.CV.min = min(mean),
              method = unique(method),
              method.full = unique(method.full),
              n_bl.covariate = unique(n_bl.covariate),
              .groups = "drop") %>%
#    mutate(method.and.n_bl.cov = paste(method, n_bl.covariate, sep = "_")) %>%
    # for label at last point
    mutate(
      label = if_else(
        prediction_time == max(prediction_time), model.name, NA_character_)) %>%
  #  filter(method == "pCox-lm" | method == "pencal" | method == "pCox-bl" | method == "pencal-min" | method == "pencal-1se") %>%
  #  filter(n_bl.covariate == 26 | n_bl.covariate == 5) %>%
    ggplot(aes(
      x = prediction_time, 
      y = repeated.CV.mean, 
      group = method,
      color = method)
    ) +
   # geom_errorbar(aes(
   #   ymax = repeated.CV.max,
   #   ymin = repeated.CV.min),
   #   alpha = 0.3,
   #   width = 0.3, size = 0.8) +
    # plot landmark time
    geom_vline(
      xintercept = t_landmark, 
      color = "grey80", linetype = "dashed", size = 1, alpha = 1
      ) + 
    geom_line(
      aes(linetype = method),
      alpha = 0.8, size = 0.8
      ) +
    geom_point(
      aes(shape = method), 
      alpha = 0.8, size = 2.5
      ) +
    # label lines, but too congested
#    geom_label_repel(
#      aes(label = label),
#      nudge_x = 1,
#      na.rm = TRUE, ) +
    scale_x_continuous(breaks = 1:15) + 
  #  scale_y_continuous(breaks = 0:10 / 10) +
#    coord_cartesian(ylim = c(0, NA)) +
    coord_cartesian(
      xlim = c(1, 15),
      ylim = c(0, 0.25)
      ) +
    ggsci::scale_color_lancet(labels = labels.method) +
    scale_linetype_discrete(labels = labels.method) +
    scale_shape_discrete(labels = labels.method) +
    xlab("Year from baseline") + ylab("Brier score") +
    labs(
#      title = g_title,
      title = paste("Landmark time =", t_landmark),
#      caption = g_caption,
  #    caption = "Error bar represents the max and min",
  #    linetype = "number of (candidate) baseline covariates",
      color = "Method",
      linetype = "Method",
      shape = "Method"
    )  +
    theme_pubr() +
    theme(
      legend.position = "right", 
      legend.box = "vertical",
      panel.grid.major.y = element_line(color = "grey60"),
      panel.grid.minor.y = element_line(color = "grey80")
      ) 
  
  return(list(
    landmark = t_landmark,
    g = g
  ))
})
```

print brier
```{r, fig.width=12, fig.height=6}
# Print
for (x in list.g.brier) {
  print(x$g)
}
```

save brier
```{r}
for (x in list.g.brier) {
  ggsave(
    filename = paste0("compare_brier_lm", x$landmark, ".png"),
    plot = x$g,
    device = "png",
    path = "./figure/compare_brier/",
    width = 12, height = 6
  )
}

```

combined plot
```{r, fig.width=8, fig.height=12}
g.brier.combined <- ggpubr::ggarrange(
  plotlist = lapply(list.g.brier, function(x) x$g), 
  ncol = 2, nrow = 3, common.legend = TRUE
)

ggsave(
  filename = "compare_brier_combined.png",
  plot = g.brier.combined,
  device = "png", 
  path = "./figure/compare_brier/",
  width = 10, height = 12,
  bg = "white"
)
```


